---
title: "Page 2"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    css: style.css
---

This is the top of the page. We want a better title/top font.

# CSS works too {#okthen .emphasized}

# Summaries {.tabset .tabset-fade .tabset-pills}

Here are some summaries of R things

Let's

make

some

space

## IO 

Take a look at this:
```{r}
x <- rnorm(100); y <- x + rnorm(100)
```

We can see that fgklj sjklkjf llkjf! 

More space

## Plotting {.tabset .tabset-fade .tabset-pills}

### Base

```{r}
plot(rnorm(10))
```

```{r}
library(datasets)
hist(airquality$Ozone)
```

```{r}
with(airquality, plot(Wind, Ozone))
```

```{r}
airquality <- transform(airquality, Month=factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
```

### Lattice

Lattice plots

### ggplot2

```{r}
library(ggplot2)
str(mpg)  # mpg comes with ggplot2
head(mpg$drv)
head(mpg$manufacturer)
qplot(displ, hwy, data = mpg)  # qplot(x, y, dataframe)
qplot(displ, hwy, data = mpg, col = drv)  # auto creates legend also
qplot(displ, hwy, data = mpg, shape = drv, col = drv) # colors and shapes per drv factor
qplot(displ, hwy, data = mpg, geom = c("point", "smooth")) # adding a statistic (geom)
qplot(displ, hwy, data = mpg, col = drv, geom = c("point", "smooth"))  # both
qplot(displ, hwy, data = mpg, col = drv) + geom_smooth(method="lm")

# split it with facets
qplot(displ, hwy, data = mpg, facets = drv ~ .) + geom_smooth(method="lm") # 3 cols
qplot(displ, hwy, data = mpg, facets = . ~ drv) + geom_smooth(method="lm") # 3 rows

qplot(hwy, data = mpg, fill = drv, geom = "density") # smooth density curve
```

qplot(hwy, data = mpg, fill = drv, geom = "density") # density and areas
qplot(hwy, data = mpg, col = drv, geom = "density")  # just the lines

## *apply Loop Functions

tapply, sapply, mapply, apply, vapply, ...

## plyr and dplyr

So useful!

# Coursera Stuff