---
title: "Fundamentals"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: yeti
    css: style.css
---

### Data Types

NA vals, logical vectors, factors

#### Factors

A factor is like a bunch of labels. “Yes”, “No”, “Maybe” for instance. In that case those three would be the Levels. The actual factor would be something like yes yes yes no maybe no no yes maybe. Factors are ordered or unordered. Yes, no, maybe is unordered but child, parent, grandparent would probably be ordered.

Here "a" is our baseline, the levels are lexicographic. 
```{r}
fac <- factor(c("a", "a", "c", "a", "c", "m", levels=c("m", "a","c"))); print(fac)
```

```{r}
fac <- relevel(fac, ref="m"); print(fac)
unclass(fac)
```

Behind the scenes, as seen above with unclass()

#### NA

A NaN is a type of NA.
```{r}
is.na(NaN)
Inf + Inf
Inf - Inf
Inf * Inf
Inf / Inf
0+0
0-0
0*0
0/0
Inf * 0
Inf + 0
Inf - 0
Inf / 0
0 / Inf
```

#### Matrix

Matrices are vectors with a dimension attribute
```{r}
v1 <- 1:6
dim(v1) <- c(2,3)
class(v1)
```

Giving dimension names to a matrix
```{r}
A <- matrix(nrow = 3, ncol = 2)
dimnames(A) <- list(c("row1", "row2", "row3"), c("col1", "col2"))
A
```

cbind and rbind
```{r}
rbind(1:3, 6:8)
cbind(1:3, 6:8)
```

### Coercion

Coerce to matrix, note the T's and F's
```{r}
x <- data.frame(id = 1:4, truth = c(T,T,F,T))
data.matrix(x)
```

```{r}
my_matrix <- matrix(1:20, 4,5)
patients <- c("Sean", "Bill", "Claude", "Kelly")
cbind(patients, my_matrix)       # coerces everything to char
data.frame(patients, my_matrix)  # patients is char, matrix is matrix
```

```{r}
as.logical(-2:2)
as.numeric(TRUE)
as.numeric(c("a","b","c"))  # no way to coerce char -> num
```

### IO, files

dir.create, file.info, file.rename, ls, dir, unlink, 

Filesystem independent path reference, nest dirs by recursive = TRUE
```{r}
file.path("folder1", "folder2")
dir.create(file.path("testdir2", "tesdtdir3"), recursive = TRUE)
unlink("testdir2", recursive = TRUE)
```

### Functions

args, dotdotdot example .R

#### Scope 

#### anonymous/closure/lambda 