---
title: "Plots"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    css: style.css
---

(Area above the first section)

# Base {.tabset .tabset-fade .tabset-pills}

## About

The base plotting system in R

### plot()

yep

yep

yep

yep

yep

yep

yep

yep


### hist()

Histogram

Histogram

Histogram

Histogram

Histogram

Histogram

Histogram

Histogram

Histogram

## Examples

```{r}
plot(rnorm(10), rnorm(10))
```

Other examples

Other examples

Other examples

Other examples

Other examples

Other examples

Other examples

Other examples


# Lattice {.tabset .tabset-fade .tabset-pills}

LATTICE

LATTICE

LATTICE

LATTICE

LATTICE

LATTICE


## About

...


...

...

...

...

...

## Examples

...

...

...

...

...

...

# ggplot2 {.tabset .tabset-fade .tabset-pills}

## About

Based on Grammar of Graphics etc

Based on Grammar of Graphics etc

Based on Grammar of Graphics etc

Based on Grammar of Graphics etc

Based on Grammar of Graphics etc

Based on Grammar of Graphics etc

Based on Grammar of Graphics etc

## Examples

```{r}
library(ggplot2)
g <- ggplot(mpg, aes(cyl, cty))
summary(g) # tells our data, mapping and faceting
print(g)   # nothing, haven't specified our geometry
g + geom_point()  # comes with a default printing method
p <- g + geom_point()  # and we can save it, too
print(p + geom_smooth())  # adding a smoother, "loess" by default
```

